Adaptive Disk Storage for Interaction Graphs

### 2014-05-15 docs/graph-adapt/main.tex
updated problem description

% 
Typical interaction graphs have data associated with their vertices and edges.
Most algorithms access this data as they traverse the graph. However, not all
of the data is accessed by all the queries. Typically, there are correlations
among the attributes accessed by different queries, such as Q1 and Q5
accessing attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and
so on.
%

In an interaction graph database organized as blocks of temporal neighborlists
(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and vertex
properties locally. For  instance, if the edge properties are kept away in a
relational table, there will be almost no benefit to the locality
optimizations performed for block organization (discussed in Gedik et al.~\cite{gedik14}, as we would have to go back and forth between the disk blocks to
access the edge attributes. 
%

@@

Recall that interaction graphs are temporal. As such, the co-access
correlations for the attributes can be different for different temporal
regions. It might also  be unknown at the insertion time and may be discovered
later depending on the workload. This points to the need for adaptively
optimizing the layout (somewhat similar to the H2O paper).
%

I imagine a solution to this problem, which I name the 'rail layout'. The idea
is to start with large blocks that contain the entire data. As we learn about
the access properties for different time regions, we might start splitting
such blocks into smaller blocks that run parallel to each other, almost like
having two or more graph databases for certain time regions, each containing a
 different subset of the attributes, but with a link between them in case a
query needs to access both. Some attributes can be replicated as well. 
More details to follow about all this...

\section{Example}

\begin{alltt}\scriptsize
- Work through a concrete example of a database and a 
   couple of queries that would change the layout on disk.
\end{alltt}

@@

@@

\section{Adaptation}

@@

%\linprog{f(x)}{g_i(x) \leq 0}{\eqnlimit[1]{i}[n]}{h_j(x) = 0}{\eqnlimit[1]{j}[m]}{eq:linprog1}

 \begin{alignat*}{2}
    \text{minimize: }   & L(\mathcal{P}, B)   \\
    \text{subject to: } & H(\mathcal{P}, B) \leq \text{Max Storage}
  \end{alignat*}

@@

>>>>

% 
Typical interaction graphs have data associated with their vertices and
edges. We call this data \emph{attributes} or \emph{properties}.
Most algorithms access this data as they traverse the graph. However, not all
of the data is accessed by all the queries. Typically, there are correlations
among the attributes accessed by different queries, such as Q1 and Q5
accessing attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and
so on.
%

In an interaction graph database organized as blocks of temporal neighborlists
(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and vertex
properties locally. For  instance, if the edge properties are kept away in a
relational table, there will be almost no benefit to the locality
optimizations performed for block organization (discussed in Gedik et al.~\cite{gedik14}, as we would have to go back and forth between the disk blocks to
access the edge attributes). 
%

@@

An important characteristic of interaction graphs are that they are temporal.
As such, the co-access correlations for the attributes can be different for
different temporal regions. It might also be unknown at the insertion time and
may be discovered later depending on the workload. This points to the need for
adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
and HYRISE~\cite{grund10}).
%

I imagine a solution to this problem, which I name the 'rail layout'. The idea
is to start with large blocks that contain the entire data. As we learn about
the access properties for different time regions, we might start splitting
such blocks into smaller blocks that run parallel to each other, almost like
having two or more graph databases for certain time regions, each containing a
 different subset of the attributes, but with a link between them in case a
query needs to access both.

\section{Problem Description}

The problem is to determine a partitioning of the disk blocks into sub-blocks
that would minimize query I/O. Of course, partitioning into sub-blocks results
in an increase in storage cost. The storage cost must stay below some specified limit.

There are two possibilities: \emph{overlapping}
and \emph{non-overlapping} partitioning. In the overlapping case, attributes can
be replicated in the sub-blocks. The non-overlapping case is a true partition.

We start by defining the cost model, which formalizes the query I/O and storage costs.

@@

\paragraph{\textbf{Other useful equations}}

@@

\section{Partitioning}

TODO: This section is still a work-in-progress. We are trying to find the algorithms.

@@

\textbf{Problem:} \emph{Find a true partitioning of attributes that minimizes
  the query I/O and bounds the storage cost by some upper limit.}

@@

\textbf{Problem:} \emph{Find an overlapping partitioning of attributes that minimizes
  the query I/O and bounds the storage cost by some upper limit.}

Below, we sketch two proposed algorithms. We are still working on these.
